\Large \textbf{ Phenotype and gene ontology enrichment as guides for disease modeling in \cel{} }

### 2017-01-17 pea_paper_docs/tex/pea.tex
added most citations to pea.

\section*{Introduction}
The last decade has seen an explosion of techniques capable of genome-wide
measurements. Some examples of genome-wide tools include RNA-seq~\cite{} to
measure gene expression, CHIP-seq~\cite{} to measure binding, or genome-wide
methylation profiling to understand epigenetic changes~\cite{}. These tools are capable
of generating large quantities of data. Understanding these data, and generating
hypotheses from them remains challenging. A common approach used to understand
these datasets is to reduce the dimensionality of the data via enrichment
analyses of ontologies~\cite{}, which helps researchers understand what terms are
overrepresented beyond random levels. By analyzing overrepresented terms in
aggregate, researchers can better understand what biological processes were most
affected in a given experiment, and form hypotheses about what is
happening~\cite{}.
This approach is limited by what ontologies can be tested for enrichment.
The best-known ontology for biological research is the Gene Ontology (GO), which
provides a controlled language to describe molecular and cellular functions of
genes~\cite{}. In \cel{}, curated tissue and phenotype ontologies exist which provide
controlled languages with which to describe \cel{} anatomy and phenotypes
respectively~\cite{}. However, enrichment tools only exist for gene and tissue ontologies
in the community today (see for example~\cite{}). Another limitation is that
tissue enrichment testing is not offered on the same websites as GO enrichment
testing, which requires users to test their data on different websites that may
or may not use different methodologies to detect enrichment.

Another way to use enrichment tools is for evolutionary comparison purposes.
In molecular biology it is often useful to know when a gene is homologous
between two species---that is to say, common by descent---because knowledge of
homology often brings with it knowledge of function~\cite{}. Indeed, many
important gene regulatory networks (GRNs) are conserved between organisms as
highly diverged as nematodes and humans (for example, see~\cite{}). While genes
and GRNs may be conserved between species, their outputs often differ however.
For example, although homeobox genes are known to be important for limb
formation in many animals~\cite{}, these genes do not form limbs in nematodes.
The concept of a phenolog has been put forward to explain relationships between
phenotypes that have the same underlying genetic regulatory network~\cite{}.
Formally, two phenotypes are phenologs of each other if the homologs of the
genes that cause a phenotype in an organism cause a second phenotype in another.

To study a clinically relevant disease in a non-human, an appropriate
model has to be established. A straightforward method towards establishing a
disease model in
\cel{} is to link a disease to a causal gene, then to identify the homologous
gene in \cel{} and then to study the function of the genetic homolog to
extrapolate back to humans. However, this method relies on the existence of
known disease genes and requires that the homolog have a phenotype that can be
reliably identified and studied. A fundamentally different way to establish a
disease model in \cel{} would be to identify the phenologs of the disease to be
studied in \cel{} by identifying disease-associated human genes in an unbiased
manner through genome-wide association studies (GWAS) and identified candidate
homolog genes in \cel{}. The homologs can be used to identify \cel{} disease
phenologs, which can in turn be used as the basis for screens to identify
genes that are associated with that phenolog. Approaches similar to this have
been successfully used in the past to make non-obvious links between phenotypes
in different species~\cite{}.

@@

An additional problem with genome-wide
queries of \cel{} states (be they developmental, neuronal, or other) is that
they do not
always have a straightforward interpretation in terms of phenotypes. In these
situations, researchers must rely on intuition to select a phenotype to screen
for. As a result, many hits may go unexplored that would prove fruitful. The
question of how to design a screen that is maximally informative is an important
question that has so far not been addressed within this community.

To facilitate understanding of large datasets, and to make discovery of
phenologs easier, we have completed an enrichment tool suite in WormBase
that allows users to rapidly perform phenotype, tissue and gene ontology
enrichment analyses (PEA, TEA and GEA respectively) on curated \cel{} ontologies
using the same methodology for each one. They are located at
% TODO: include links.
We applied our tools towards the unbiased discovery of phenologues of
multigenic, complex diseases including
% Diseases to study:
systemic lupus erythematosus, obesity and obesity-related traits,  and
rheumatoid arthritis
by using genes associated with these
diseases via genome-wide association studies. We also illustrate the utility of
the complete enrichment suite for finding new relationships in complex data by
analyzing a ciliary neuron transcriptome~\cite{}.

\section*{Methods}
\subsection*{Human disease phenolog identification}
We used the GWAS EBI-NHGRI catalog~\cite{} to extract information on all genome-wide
association studies deposited there. We only selected traits that had $>300$
associated genes. We identified 24 traits that met our criteria. Next, we used
DIOPT~\cite{} to identify candidate homologs for the genes associated with these
traits. Briefly, DIOPT combines a large number of methods for identifying homologs
and returns homolog candidates associated with a compound score. Depending on the score,
homologs can be considered `high', `moderate' or `low' rank, reflecting confidence
in the homology. Many-to-one and one-to-many homology relationships are allowed
in DIOPT, reflecting a mixture of uncertainty and family expansion/reduction.
For our study, we only accepted homolog candidates with `high' or `moderate' scores
and we did not insist on a one-to-one relationship between genes.

@@

After we identified worm homologs for each trait, we reassessed how many traits
still had $>100$ gene candidates, and dropped all traits that had less than this
for our analysis. We identified 18 traits that met this criteria. The gene
lists for each of these 18 traits were then analyzed for gene, tissue and
phenotype enrichment (see~Fig.\ref{fig:gwas}). Tissue enrichment was
performed using the WormBase Tissue Enrichment Analysis (TEA) tool~\cite{}.

\section*{Results}
\subsection*{Developing the WormBase enrichment suite}
% TODO add numbers instead of XXX
We developed the dictionaries for PEA and GEA using the same procedure as was
used for TEA~\cite{}. We generated a dictionary that included terms with at
least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
(the total number of terms in the dictionary was 251, annotated by 9,169 genes
 for the version XXXXX);\@ and we generated a
dictionary that included terms with at least 50 annotating genes or more and
had a similarity threshold of 0.95 for GEA (the total number of terms in this
dictionary is 271, annotated by 14,636 genes for the version XXXX).
\@ Next, we benchmarked the dictionaries on the same gene
sets as TEA and obtained enrichment of all the expected categories. For example,
on a gene set enriched for embryonic muscle genes~\cite{}, the top two enriched
phenotype terms by q-value were `muscle system morphology variant' and `body
wall muscle thick filament variant'; the top two enriched GO terms were
`myofibril' and `striated muscle dense body'. For all the benchmarking
results, see supplementary information. Having generated and validated our
dictionaries, we proceeded to identify phenologs for several common human
diseases.

\subsection*{Applying the WormBase enrichment suite}
To discover phenologs, we first needed to identify genes that
contribute to a disease in an unbiased manner. One way to discover gene
associations in an unbiased manner is to perform GWSA in human populations.
Therefore, we used the GWAS NHGRI-EBI
Catalog~\cite{} to identify genes associated with human diseases. We found the
best nematode candidate homologs for these genes using DIOPT~\cite{} and applied
our enrichment suite to each of these gene regulatory networks.

@@

\subsection*{Systemic lupus erythematosus}
Systemic lupus is an autoimmune disease that is believed to be polygenic in
nature~\cite{}. It mainly affects women and is characterized by painful
and swollen joints, hair loss, and fatigue~\cite{}. Since worms do not have a
cellular immune system, we were interested in what phenologs corresponded to
this disorder in \cel{}. To establish phenolog candidates, we obtained
\hlupus{} genes associated with the disease via GWAS studies, and found
\wlupus{} homolog candidates in \cel{}.

@@

\subsection*{Rheumatoid arthritis}
Rheumatoid arthritis is an auto-immune disease that is characterized by
swollen and painful joints that progressively deteriorate~\cite{}. Unlike lupus,
rheumatoid arthritis is not life-threatening~\cite{}, and comorbidity between rheumatoid
arthritis and lupus is low~\cite{}, suggesting that they may have at least
partially distinct genetic causes. We found \harthritis{} genes associated with
rheumatoid arthritis, for which we found \warthritis{} worm homolog candidates.

% TODO: Is no tissue enrichment observed bc poor annotations, or because
% all tissues randomly represented?
The only phenotype that was enriched for these homologs was `short' (10 genes,
\qval{4}), even though 64 homologs were associated with at least
one phenotype term. No tissue was enriched in this dataset. Because
82 genes are annotated to have expression in at least one tissue, the lack of
enrichment does not reflect ignorance about the sites of expression of these
genes. GEA showed that enriched
molecular functions for these genes include `collagen trimer' (22 genes,
\qval{15}). However, this term was enriched as the result of degeneracy in the
homolog candidates for the SFTPD gene. Other terms included
 `glycosylation' (10 genes, \qval{4}) and `Golgi apparatus' (11
genes, \qval{3}), but this enrichment was also the result of degenerate homolog
candidates for the human gene B3GNT7 which encodes a
beta-1,3-N-acetylgalactosaminyltransferase.

The `short' phenotype was the result of the \emph{cat-4}, \emph{dpy-7},
\emph{rnt-1}, \emph{sem-4}, \emph{unc-116}, \emph{ocrl-1} and some genes in the
\emph{fat} family. Although these genes are bound by a common phenotype, any
genetic relationships between these genes are not immediately clear. Some genes,
like \emph{sem-4} and \emph{rnt-1} are likely transcription factors with roles
in development (including hypodermal development).
Others are molecular motors (\emph{unc-116}) that are broadly expressed throughout
the body of \cel{}. Yet others have known roles in neuron and muscle function,
such as \emph{ocrl-1} and \emph{cat-4}. The `short' phenotype is a subset of
the `body length variant' phenotype. Body length in \cel{} can be controlled via
cell size, shape and number\cite{}; alternatively, cuticle development can alter body
shape~\cite{}; finally, muscles can alter the effective body length artificially~\cite{}.

@@

\subsubsection*{Enrichment in the ciliary neuronal transcriptome}
As an example of how ontology enrichment can improve our understanding of
transcriptomes,
we selected a ciliary neuron dataset~\cite{} and ran the complete WormBase
Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
they are also present in the male cephalic sensillum and hermaphrodites also have
ciliated neurons. PEA reveals that the ciliated
neuron transcriptome is enriched for genes that are typically associated with
`meiotic chromosome segregation' (46 genes, \qval{5}),
`aneuploidy' (42 genes, \qval{5}) and `spindle defective early embryos' (45
genes, \qval{2}) (see Fig.~\ref{fig:cilia}).

@@

Although one interpretation of the results would be that microtubule genes
are driving the enrichment of these terms, another possibility is that
there are cell-cycle genes that are driving the enrichment of these phenotypes
and tissues. Indeed, GO enrichment shows terms such as `DNA replication' (29
genes, \qval{5}), and `purine NTP-dependent helicase activity'
(15 genes, \qval{1}). Visual inspection of list in question reveals that
cell-cycle and DNA replication/repair genes are abundant in this transcriptome
and include genes such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This
analysis reveals that the ciliary neuron transcriptome is enriched in
genes associated with microtubules, but the cell-cycle machinery is also
carefully regulated.

\subsection*{Deconstructing phenotype-tissue relationships}
\subsubsection*{Tissue enrichment on the Egl gene set reveals cellular
components of the phenotype}
How does a phenotype emerge? We realized that with the tools that we have
developed, it is possible to understand what tissues contribute to a phenotype
in a probabilistic framework. In other words, we can extract all genes associated
with a particular phenotype, then search for tissue terms that are enriched to
understand how a phenotype arises from interactions between anatomical regions.
As a test of this, we selected the egg-laying defective (Egl) phenotype.
In \cel{}, egg-laying is a complex behavior that involves
a large number of tissues~\cite{}. The somatic gonad acts as a repository for the eggs,
the uterine seam cells help protect the uterus, and a variety of muscles help
contract the uterus and open the vulva to lay an egg~\cite{}. The vulva must be well-formed
to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
in the egg-laying control~\cite{}. The complexity of the interactions that
happen to allow egg-laying make understanding the Egl phenotype in terms of
tissues a challenging task.

@@

\section*{Conclusions}

We have highlighted three possible uses for phenotype and GO enrichment analyses.
Taken together, the WormBase Enrichment Suite can help guide researchers as they
search for a phenologue that is representative of a human disease that has no
immediately obvious counterpart in the worm. Such phenologues may benefit from
the fact that they represent an unbiased approach to disease modeling as long as
the human genes are selected from unbiased screens. GWAS data suggests that certain
aspects of lupus biology may be best understood from a developmental perspective
in \cel{}. On the other hand, our results suggest that rheumatoid arthritis
might not be effectively modeled in the worm, although it is possible that our
inability to find phenologues more concretely is the result of the GWAS screens:
If GWAS screens identified factors that lead to worse prognosis of rheumatoid
arthritis, but not factors that cause onset of the disease, this could explain
the lack of phenotype enrichment in \cel{}.

% Phenologues

% Screen Design

% Curation and annotation biases

The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
an important step towards a unified set of analyses that can help researchers
to understand genomic datasets. By offering tissue, phenotype and gene ontology
enrichment on a single site, researchers will benefit both in the speed at which
the analysis is completed and from the fact that the analyses are all
carried out with the same methodology and the same underlying algorithms.

>>>>

\section*{Introduction}
The last decade has seen an explosion of techniques capable of genome-wide
measurements. Some examples of genome-wide tools include
RNA-seq~\cite{Mortazavi2008} to measure gene expression or
CHIP-seq~\cite{Johnson2007} to measure protein binding. These tools are capable
of generating large quantities of data. Understanding these data, and generating
hypotheses from them remains challenging. A common approach used to understand
these datasets is to reduce the dimensionality of the data via enrichment
analyses of ontologies~\cite{TheGeneOntologyConsortium2000a}, which helps
researchers understand what terms are
overrepresented beyond random levels. By analyzing overrepresented terms in
aggregate, researchers can better understand what biological processes were most
affected in a given experiment, and form hypotheses about what is
happening~\cite{Rhee2008}.
This approach is limited by what ontologies can be tested for enrichment.
The best-known ontology for biological research is the Gene Ontology (GO), which
provides a controlled language to describe molecular and cellular functions of
genes~\cite{TheGeneOntologyConsortium2000a}.
In \cel{}, gene, tissue and phenotype ontologies exist
% which provide
% controlled languages
with which to describe \cel{} anatomy and phenotypes
respectively~\cite{Schindelman2011,Lee2003}. These ontologies are curated by
professional curators at WormBase, which is a repository of all \cel{}
data~\cite{Howe2016}.
However, enrichment tools
only exist for gene and tissue ontologies
in the community today (see for example~\cite{Chikina2009,Mi2013,Angeles-Albores2016}).
Another limitation is that
tissue enrichment testing is not offered on the same websites as GO enrichment
testing, which requires users to test their data on different websites that may
or may not use different methodologies to detect enrichment.

Another way to use enrichment tools is for evolutionary comparison purposes.
In molecular biology it is often useful to know when a gene is homologous
between two species---that is to say, common by descent---because knowledge of
homology often brings with it knowledge of function. Indeed, many
important gene regulatory networks (GRNs) are conserved between organisms as
highly diverged as nematodes and humans (for example, see~\cite{Sternberg1998}).
While genes
and GRNs may be conserved between species, their outputs often differ however.
For example, although homeobox genes are known to be important for limb
formation in many animals~\cite{Zakany2007}, these genes do not form limbs in nematodes.
The concept of a phenolog has been put forward to explain relationships between
phenotypes that have the same underlying genetic
regulatory network~\cite{McGary2010,Lehner2013}.
Formally, two phenotypes are phenologs of each other if the homologs of the
genes that cause a phenotype in an organism cause a second phenotype in another.

To study a clinically relevant disease in a non-human, an appropriate
model has to be established. A straightforward method towards establishing a
disease model in
\cel{} is to link a disease to a causal gene, then to identify the homologous
gene in \cel{} and then to study the function of the genetic homolog to
extrapolate back to humans. However, this method relies on the existence of
known disease genes and requires that the homolog have a phenotype that can be
reliably identified and studied. A fundamentally different way to establish a
disease model in \cel{} would be to identify the phenologs of the disease to be
studied in \cel{} by identifying disease-associated human genes in an unbiased
manner through genome-wide association studies (GWAS) and identified candidate
homolog genes in \cel{}. The homologs can be used to identify \cel{} disease
phenologs, which can in turn be used as the basis for screens to identify
genes that are associated with that phenolog. Approaches similar to this have
been successfully used in the past to make non-obvious links between phenotypes
in different species~\cite{McGary2010}.

@@

An additional problem with genome-wide
queries of \cel{} states (be they developmental, such as
L1, L2, dauer; behavioral states such as awake versus asleep; or other) is that
they do not
always have a straightforward interpretation in terms of phenotypes. In these
situations, researchers must rely on intuition to select a phenotype to screen
for. As a result, many hits may go unexplored that would prove fruitful. The
question of how to design a screen that is maximally informative is an important
question that has so far not been addressed within this community.

To facilitate understanding of large datasets, and to make discovery of
phenologs easier, we have completed an enrichment tool suite in WormBase
that allows users to rapidly perform phenotype, tissue and gene ontology
enrichment analyses (PEA, TEA and GEA respectively) on curated \cel{} ontologies
using the same methodology for each one. They are located at
% TODO: include links.
We applied our tools towards the unbiased discovery of phenologues of
multigenic, complex diseases including
% Diseases to study:
systemic lupus erythematosus, obesity and obesity-related traits,  and
rheumatoid arthritis
by using genes associated with these
diseases via genome-wide association studies. We illustrate the utility of
the complete enrichment suite for finding new relationships in complex data by
analyzing a ciliary neuron transcriptome~\cite{Wang2015}. Finally, we show that
the dictionaries generated for these enrichment analyses can help elucidate the
contributions of specific tissues to specific phenotypes.

\section*{Methods}
\subsection*{Implementing the enrichment analyses}
All scripts were implemented in Python 3.5~\cite{Rossum2011}. We used
pandas~\cite{McKinney2011} and scipy~\cite{Oliphant2007} to write the
statistical testing framework. Matplotlib~\cite{Hunter2007} and Seaborn~\cite{Waskom}
libraries are used to generate all plots. Testing was performed using the
WormBase version XXXXX. The WormBase Enrichment suite can be installed using
pip via the command:

\texttt{pip install tissue\_enrichment\_analysis}

A brief tutorial on how to use the command line tool can be found in the SI.
A brief tutorial on how to interact with the suite using python is available
as a Jupyter notebook~\cite{Perez2007} in the SI.

\subsection*{Human disease phenolog identification}
We used the GWAS EBI-NHGRI catalog~\cite{MacArthur2016} to extract information
on all genome-wide
association studies deposited there. We only selected traits that had $>300$
associated genes. We identified 24 traits that met our criteria. Next, we used
DIOPT~\cite{Hu2011} to identify candidate homologs for the genes associated with these
traits. Briefly, DIOPT combines a large number of methods for identifying homologs
and returns homolog candidates associated with a compound score. Depending on the score,
homologs can be considered `high', `moderate' or `low' rank, reflecting confidence
in the homology. Many-to-one and one-to-many homology relationships are allowed
in DIOPT, reflecting a mixture of uncertainty and family expansion/reduction.
For our study, we only accepted homolog candidates with `high' or `moderate' scores
and we did not insist on a one-to-one relationship between genes.

@@

After we identified worm homologs for each trait, we reassessed how many traits
still had $>100$ gene candidates, and dropped all traits that had less than this
for our analysis. We identified 18 traits that met this criteria. The gene
lists for each of these 18 traits were then analyzed for gene, tissue and
phenotype enrichment (see~Fig.\ref{fig:gwas}). Tissue enrichment was
performed using the WormBase Tissue Enrichment Analysis (TEA)
tool~\cite{Angeles-Albores2016}.

\section*{Results}
\subsection*{Developing the WormBase enrichment suite}
% TODO add numbers instead of XXX
We developed the dictionaries for PEA and GEA using the same procedure as was
used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that
included terms with at
least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
(the total number of terms in the dictionary was 251, annotated by 9,169 genes
 for the version XXXXX);\@ and we generated a
dictionary that included terms with at least 50 annotating genes or more and
had a similarity threshold of 0.95 for GEA (the total number of terms in this
dictionary is 271, annotated by 14,636 genes for the version XXXX).
\@ Next, we benchmarked the dictionaries on the same gene
sets as TEA and obtained enrichment of all the expected
categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
Pauli2006, Portman2004, Fox2007, Smith2010}.
For example,
on a gene set enriched for embryonic muscle genes~\cite{Watson2008a},
the top two enriched
phenotype terms by q-value were `muscle system morphology variant' and `body
wall muscle thick filament variant'; the top two enriched GO terms were
`myofibril' and `striated muscle dense body'. For all the benchmarking
results, see supplementary information. Having generated and validated our
dictionaries, we proceeded to identify phenologs for several common human
diseases.

\subsection*{Applying the WormBase enrichment suite}
To discover phenologs, we first needed to identify genes that
contribute to a disease in an unbiased manner. One way to discover gene
associations in an unbiased manner is to perform GWSA in human populations.
Therefore, we used the GWAS NHGRI-EBI
Catalog~\cite{MacArthur2016} to identify genes associated with human diseases. We found the
best nematode candidate homologs for these genes using DIOPT~\cite{Hu2011} and applied
our enrichment suite to each of these gene regulatory networks.

@@

\subsection*{Systemic lupus erythematosus}
Systemic lupus is an autoimmune disease that is believed to be polygenic in
nature~\cite{Mohan2015}. It mainly affects women and is characterized by painful
and swollen joints, hair loss, and fatigue~\cite{Lisnevskaia2014}. Since worms do not have a
cellular immune system, we were interested in what phenologs corresponded to
this disorder in \cel{}. To establish phenolog candidates, we obtained
\hlupus{} genes associated with the disease via GWAS studies, and found
\wlupus{} homolog candidates in \cel{}.

@@

\subsection*{Rheumatoid arthritis}
Rheumatoid arthritis is an auto-immune disease that is characterized by
swollen and painful joints that progressively deteriorate~\cite{Smolen2016}. Unlike lupus,
rheumatoid arthritis is not life-threatening, and comorbidity between rheumatoid
arthritis and lupus has not been described in past comorbidity studies~\cite{Dougados2013},
suggesting that they may have at least
partially distinct genetic causes. We found \harthritis{} genes associated with
rheumatoid arthritis, for which we found \warthritis{} worm homolog candidates.

% TODO: Is no tissue enrichment observed bc poor annotations, or because
% all tissues randomly represented?
The only phenotype that was enriched for these homologs was `short' (10 genes,
\qval{4}), even though 64 homologs were associated with at least
one phenotype term. No tissue was enriched in this dataset. Because
82 genes are annotated to have expression in at least one tissue, the lack of
enrichment does not reflect ignorance about the sites of expression of these
genes. GEA showed that enriched
molecular functions for these genes include `collagen trimer' (22 genes,
\qval{15}). However, this term was enriched as the result of degeneracy in the
homolog candidates for the SFTPD gene. Other terms included
 `glycosylation' (10 genes, \qval{4}) and `Golgi apparatus' (11
genes, \qval{3}), but these terms were enriched as the result of degenerate homolog
candidates for the human gene B3GNT7 which encodes a
beta-1,3-N-acetylgalactosaminyltransferase.

The `short' phenotype was the result of the \emph{cat-4}, \emph{dpy-7},
\emph{rnt-1}, \emph{sem-4}, \emph{unc-116}, \emph{ocrl-1} and some genes in the
\emph{fat} family. Although these genes are bound by a common phenotype, any
genetic relationships between these genes are not immediately clear. Some genes,
like \emph{sem-4} and \emph{rnt-1} are likely transcription factors with roles
in development (including hypodermal development)~\cite{Desai1988,Ji2004}.
Others are molecular motors (\emph{unc-116}) that are broadly expressed throughout
the body of \cel{}. Yet others have known roles in neuron and muscle function,
such as \emph{ocrl-1} and \emph{cat-4}. The `short' phenotype is a subset of
the `body length variant' phenotype. Body length in \cel{} can be controlled via
cell size, shape and number\cite{}; alternatively, cuticle development can alter body
shape~\cite{}; finally, muscles can alter the effective body length artificially~\cite{}.

@@

\subsubsection*{Enrichment in the ciliary neuronal transcriptome}
As an example of how ontology enrichment can improve our understanding of
transcriptomes,
we selected a ciliary neuron dataset~\cite{Wang2015} and ran the complete WormBase
Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
they are also present in the male cephalic sensillum and hermaphrodites also have
ciliated neurons. PEA reveals that the ciliated
neuron transcriptome is enriched for genes that are typically associated with
`meiotic chromosome segregation' (46 genes, \qval{5}),
`aneuploidy' (42 genes, \qval{5}) and `spindle defective early embryos' (45
genes, \qval{2}) (see Fig.~\ref{fig:cilia}).

@@

Although one interpretation of the results would be that microtubule genes
are driving the enrichment of these terms, another possibility is that
there are cell-cycle genes that are driving the enrichment of these phenotypes
and tissues. Indeed, GO enrichment shows terms such as `DNA replication' (29
genes, \qval{5}), and `purine NTP-dependent helicase activity'
(15 genes, \qval{1}). Visual inspection of list in question reveals that
cell-cycle and DNA replication/repair genes are abundant in this transcriptome
and include genes such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This
analysis reveals that the ciliary neuron transcriptome is enriched in
genes associated with microtubules, but includes genes that are thought to
interact with DNA either via repair mechanisms or cell-cycle
control~\cite{Hofmann2000,Lee2003a,Bailly2010}.

\subsection*{Deconstructing phenotype-tissue relationships}
\subsubsection*{Tissue enrichment on the Egl gene set reveals cellular
components of the phenotype}
How does a phenotype emerge? We realized that with the tools that we have
developed, it is possible to understand what tissues contribute to a phenotype
in a probabilistic framework. In other words, we can extract all genes associated
with a particular phenotype, then search for tissue terms that are enriched to
understand how a phenotype arises from interactions between anatomical regions.
As a test of this, we selected the egg-laying defective (Egl) phenotype.
In \cel{}, egg-laying is a complex behavior that involves
a large number of tissues~\cite{Li1990}. The somatic gonad acts as a
repository for the eggs,
the uterine seam cells help protect the uterus, and a variety of muscles help
contract the uterus and open the vulva to lay an egg~\cite{Sulston1977}. The
vulva must be well-formed
to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
in the egg-laying control~\cite{Schafer2005}. The complexity of the interactions that
happen to allow egg-laying make understanding the Egl phenotype in terms of
tissues a challenging task.

@@

\section*{Conclusions}
The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
an important step towards a unified set of analyses that can help researchers
to understand genomic datasets. These enrichment analyses will allow the community to
fully benefit from the data curation ongoing at WormBase. By using the same
algorithms to generate enrichment dictionaries for testing and using the
same model to test for term enrichment, these tools provide a coherent framework
with which to analyze genomic data. In particular, it is our hope that
phenotype enrichment will be of use to geneticists performing genome-wide
analyses, because they are familiar with the ontological terms that are tested
and with their biological meaning. Ideally, PEA could allow researchers to
design better screens to maximize target gene identification by quantifying
the phenotypes that are most overrepresented. An intriguing new direction of
research would be to create a controlled language for screening methods. With
such a language, we should be able to computationally suggest to a researcher
what screens one may wish to perform given a dataset.

\bibliography{pea-citations}
\bibliographystyle{pnas2011}
